Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-07-28 ms/main.tex
merge from master
compete revision of result1-2 and appendix A; revised intro

% Title page
\title{Cultural Transmission Can Explain the Evolution of Cooperative Behavior}

@@

\affil[1]{School of Computer Science, Inetdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

%%% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology.
Three major theories have been proposed to explain the evolution of cooperative behavior:

1. \emph{Kin selection} theory posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton,  an allele that determines altruistic behavior should increase in frequency when  the reproductive cost to the actor that cooperates, $c$, is less than the benefit to the recipient, $b$, times the genetic relatedness between the recipient and the actor, $r$.
This is also known as Hamilton's Rule:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

There is an ongoing debate about the extent to which kin selection explains the evolution of cooperation and altruism.
For example, it has been suggested that it can explain the cooperative behavior of worker castes of eusocial insects like the honey bee. % TODO REF
The most significant argument against kin selection is that in some cases cooperation among unrelated individuals appears to have evolved~\citep{wilson2005kin}. According to \citet{wilson2005kin}, this makes Hamilton's rule incomplete, although \citet{foster2006kin} reject this claim and argue that altruism cannot evolve without relatedness. They refer  to Hamilton, who claimed that relatedness can arise without recent common ancestry. %MWF: How?
\citet{wilson2005kin} also criticizes kin selection on the grounds that environmental or ecological factors are probably more important than relatedness in determining social actions. However, \citet{foster2006kin} argue that kin selection does not ignore ecology, and Hamilton's rule suggests that environmental factors causing a high benefit-to-cost ratio will favor cooperation.

2. \emph{Reciprocity} entails that repeated interactions or individual recognition are key components of the evolution of cooperation. In \emph{direct reciprocity} there are repeated encounters between the same two individuals, and at every encounter each individual has a choice between cooperation and defection. Hence, it may eventually pay off to cooperate if it may cause your partner to cooperate in the future.
This game-theoretic framework, known as the \emph{repeated prisoner's dilemma}, 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
can only lead to the evolution of cooperation if the cost is less than the benefit $b$ times the probability of another encounter between the same two individuals, $w$, 
%TODO Referencess
\begin{equation} \label{eq:reciprocity}
c < b \cdot w.
\end{equation}

3. \emph{Group selection} theory posits that cooperation is favored because it imparts an advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection divides the population into groups in which there are cooperators that help  other group members and defectors that do not.  % TODO add reference
Individuals reproduce proportionally to their fitness, and offspring are added to the same group as their parents.
If a group reaches a certain size it can split to two groups, so groups that grow faster will split more often.
Groups with cooperators grow faster than groups without cooperators, and
cooperation can evolve in this model when the cost $c$ is less than the benefit $b$ times the ratio between the  the number of groups $m$ and the sum of $m$ and the maximum group size $n$,
\begin{equation} \label{eq:groupselection}
c < b \cdot \frac{m}{m+n} .
\end{equation}

@@

% Models
\section*{Models}

@@

Horizontal cultural transmission occurs between peers. 
It may occur between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
Horizontal transmission is not always successful, as one peer may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\textbf{\autoref{table:interactions}}).

@@

%%% Results
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Without Oblique Transmission}

With only vertical and horizontal transmission, i.e. $v=1$, \autoref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and \autoref{eq:nextgen_parents_simplified} for the change in frequency $p'$ of phenotype $A$ among parents can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}

We find the following result and corollaries.

\textbf{Result 1: Vertical and horizontal transmission of cooperation.}
From \autoref{eq:nextgen_parents_vertical_only}, if
\begin{equation} \label{eq:unequal_transmission}
c \cdot (1-T_B) < b \cdot  \alpha T_A  + (T_A - T_B) [1 + b\tilde{p}(1-\alpha)] ,
\end{equation}
then $\tilde{p}' > \tilde{p}$, and the frequency of the cooperator phenotype $A$ among parents increases every generation.

\textbf{Corollary 1.1: Symmetric horizontal transmission.}
If $T=T_A=T_B$, then
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T},
\end{equation}
which can be seen as a cultural version of \emph{Hamilton's rule} (\autoref{eq:hamilton_rule}), where $r_H=\alpha T/(1-T)$ can be thought of as \emph{horizontal relatedness}.
Therefore, if the cost $c$ is less than the benefit $b$ times the horizontal relatedness $r_H$, then cooperation will take over of the population (see~\textbf{\autoref{fig:results_a}}).

\textbf{Corollary 1.2: Complete correlation between transmission and cooperation.}

In this case $\alpha=1$, and horizontal transmission can only occur as a result of cooperative interactions.
Therefore, \autoref{eq:unequal_transmission}, which determines the conditions for evolution of cooperation, becomes
\begin{equation}\label{eq:horizontal_transmission_only_after_interaction}
c \cdot (1-T_B) < b \cdot T_A + (T_A - T_B).
\end{equation}
This is equivalent to a result by~\citet[eq.~1]{lewin2017microbes}.

\autoref{eq:horizontal_transmission_only_after_interaction} can be written as
\begin{equation} \label{eq:horizontal_transmission_only_after_interaction2}
1 - (1-T_B)(1-c) < T_A \cdot (1+b) ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
Consider an interaction between two individuals: a cooperator and a defector.
$(1-T_B)(1-c)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-T_B)(1-c)$ is the probability that either the cooperator becomes a defector, \emph{or} that it fails to reproduce.
This is the effective cost for cooperation from this interaction.
$T_A\cdot(1+b)$ is the probability that the defector becomes cooperative and reproduces.
This is the effective benefit for cooperation from this interaction.
So, \autoref{eq:horizontal_transmission_only_after_interaction2} means that cooperation can evolve if the effective cost is less than the effective benefit.

\textbf{Corollary 1.3: No correlation between transmission and cooperation.}

In this case $\alpha=0$, and horizontal transmission is entirely independent from cooperative interactions.  
Then, \autoref{eq:unequal_transmission} becomes
\begin{equation} \label{eq:horizontal_transmission_from_population}
c\cdot(1-T_B) < b \cdot \tilde{p} \;(T_A-T_B) + (T_A-T_B). 
\end{equation}

Therefore, since all the parameters are  positive, cooperation cannot take over the population (and furthermore will become extinct) if cooperators do not have a horizontal transmission bias, i.e. if $T_A\leq T_B$.
When such a bias does exist, $T_A > T_B$, then cooperation will evolve if $\tilde{p} > \tilde{p}^{*}$, where (see also ~\textbf{\autoref{fig:results_b}})
\begin{equation} 
\begin{split} \label{eq:horizontal_transmission_from_population_equilibrium}
\tilde{p}^{*} & = \frac{c}{b} \cdot \frac{1-T_B}{T_A-T_B} - \frac{1}{b}.
\end{split}
\end{equation}
A sufficient condition for evolution of cooperation is that $\tilde{p}^{*} < 0$, namely
\begin{equation} \label{eq:sufficient_horizontal_transmission_from_population}
c < \frac{T_A-T_B}{1-T_B} .  
\end{equation}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Without Vertical Transmission}

@@

\textbf{Result 2: Oblique and horizontal transmission of cooperation.} 
If there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).

@@

\textbf{Result 3: Oblique and vertical transmission of cooperation.}
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of the cooperative phenotype $A$}, 
\begin{enumerate}
\item if $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$, or
\item if $T_A>T_B$ and $0<\beta_3$, or 
\item if $T_A<T_B$ and $\beta_1<\beta_3$.
\end{enumerate}

@@

\end{enumerate}

@@

% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation.
We studied fully mixed and very large populations with a prisoner's dilemma payoff. 
We found that under horizontal and vertical cultural transmissions, if \autoref{eq:unequal_transmission} is satisfied, cooperation will take over fully mixed populations (Result 1).
Under oblique and horizontal transmission, horizontal transmission bias for the cooperative phenotype is sufficient and necessary for evolution of cooperation (Result 2, \autoref{eq:oblique_only_result}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are further complicated. Yet, we find that cooperation can evolve and in some cases be maintained together with defection (Result 4). 
Importantly, our results demonstrate that cooperation can evolve even in a fully mixed population (i.e. in an unstructured population), without repeating interactions or individual recognition.
These results significantly further our understating of the cultural evolution of cooperation.

@@

\citet{Eshel1982} have studied a relevant model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters. This assortment may be due, for example, to population structure or active partner choice.
In this model, cooperative behavior can evolve if \citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:assortative_meeting}
c < b \cdot m,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation, and $m$ is the fraction of the population that meets an individual of the same phenotype rather than a random individual\footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
}. 
As a version of Hamilton's rule, $m$ can be interpreted as the effective relatedness.
In our Corollary 1.1 (\autoref{eq:equal_transmission}), cooperation evolves if $c < b \cdot \alpha T / (1-T)$. So in our model $\alpha T/(1-T)$ is the effective relatedness, which is affected by $\alpha$, the correlation between transmission and interaction, and $T$, the horizontal transmission rate.

@@

\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

% Bibliography
\bibliographystyle{plainnat}
\bibliography{bib}

% Appendices
\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

\section{} \label{sec:appendixA}

We start with \autoref{eq:nextgen_parents_vertical_only} and we want to investigate when $\tilde{p}< \tilde{p}'$, that is, when 
\begin{equation} \label{eq:horizontal_transmission_from_interaction_step0}
\begin{aligned}
\bar{w}\tilde{p} < & \, \tilde{p}^2 (1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
& + \tilde{p}(1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
& + \tilde{p}(1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}
First divide by $\tilde{p}$ to obtain
\begin{equation} \label{eq:horizontal_transmission_from_interaction_step1} 
\begin{aligned} 
  \bar{w} < & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
We know that the mean fitness $\bar{w} = 1 + \tilde{p}(b-c)$, thus \autoref{eq:horizontal_transmission_from_interaction_step1} becomes
\begin{equation} \label{eq:horizontal_transmission_from_interaction_step2} 
\begin{aligned} 
  1 + \tilde{p}(b-c) < & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
which we simplify to inequality~\autoref{eq:unequal_transmission}.

@@

>>>>

% Title page
\title{Cultural Transmission Can Explain the Evolution of Cooperation}

@@

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can harm an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, for example rats~\citep{rice1962altruism} and birds~\citep{krams2008experimental}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology.

\emph{Kin selection} theory posits that natural selection can favor cooperation between related individuals. The importance of relatedness to the evolution of cooperation and altruism was shown by~\citet{hamilton1964genetical}. According to Hamilton, for an allele that determines cooperative behavior to increase in frequency, the reproductive cost to the actor that cooperates, $c$, must be less than the benefit to the recipient, $b$, times the `relatedness' between the recipient and the actor, $r$. This `relatedness' coefficient $r$ measures the correlation between the gene in the actor and the gene in the recipient.
This condition is also known as Hamilton's rule:
\begin{equation} \label{eq:hamilton_rule}
c < b \cdot r.
\end{equation}

\citet{Eshel1982} have studied a relevant model for the evolution of cooperative behavior under vertical transmission. Their model included \emph{assortative meeting}, or non-random encounters. That is, if a fraction $m$ of the population interacts with an individual of the same phenotype, and $1-m$ interacts randomly. 
Such assortative meeting may be due, for example, to population structure or active partner choice.
In their model, cooperative behavior can evolve if \footnote{In an extended model, which allows an individual to encounter $N$ individuals before choosing a partner, the righthand side is multiplied by $E[N]$, the expected number of encounters \citep[eq.~4.6]{Eshel1982}.
}. 
\citep[eq.~3.2]{Eshel1982}
\begin{equation} \label{eq:assortative_meeting}
c < b \cdot m,
\end{equation}
where $b$ and $c$ are the benefit and cost of cooperation
Here, $m$ takes the role of the relatedness $r$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Models
\section*{Models}

@@

Horizontal cultural transmission occurs between pairs of individuals from the same generation. 
It occurs between social partners with probability $\alpha$, or between a random pair with probability $1-\alpha$ (see~\textbf{\autoref{fig:horizontal}}).
The assortment parameter $\alpha$ is therefore the fraction of population that receives (horizontal transmission) from the social interaction partner, and $1-\alpha$ receives randomly.
Horizontal transmission is not always successful, as one partner may reject the other's phenotype. The probability for successful horizontal transmission of phenotypes $A$ and $B$ are $T_A$ and $T_B$, respectively (\textbf{\autoref{table:interactions}}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Results
\section*{Results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Oblique and Horizontal Transmission}

@@

\begin{result}[Oblique and horizontal transmission of cooperation]
If there is a horizontal transmission bias in favor of cooperation, namely
\begin{equation} \label{eq:oblique_only_result}
T_A > T_B, 
\end{equation}
then $p'>p$, and the frequency of the cooperator phenotype among adults increases every generation.
Therefore, cooperation will evolve if the cooperator phenotype has a horizontal transmission bias (see~\textbf{\autoref{fig:results_c}}).
\end{result}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Vertical and Horizontal Transmission}

With only vertical and horizontal transmission, i.e. $v=1$, \autoref{eq:horizontal} becomes
$\hat{p} =  \tilde{p}$,
and \autoref{eq:nextgen_parents_simplified} for the frequency of the cooperative phenotype among parents in the next generation $\tilde{p}'$ can be written as
\begin{equation} \label{eq:nextgen_parents_vertical_only} 
\begin{aligned}
\bar{w} \tilde{p}' 
& = \tilde{p}^2 (1+b-c) [1 - (1-\tilde{p}) (1-\alpha) T_B] \\
& + \tilde{p}(1-\tilde{p}) (1-c) [\tilde{p} (1-\alpha) T_B + 1 - T_B] \\
& + \tilde{p}(1-\tilde{p}) (1+b) [\tilde{p} (1-\alpha) + \alpha] T_A \\
& + (1-\tilde{p})^2 \tilde{p} (1-\alpha) T_A .
\end{aligned}
\end{equation}

The fixation of either cooperation or defection, 
$\tilde{p}=0$ and $ \tilde{p}=1$, are equilibria of \autoref{eq:nextgen_parents_vertical_only}, that is, they solve $\tilde{p}'= \tilde{p}$.
We therefore assume for the remainder of the analysis that $0<\tilde{p}<1$.

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
p(1-p)\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.

Therefore, for cooperation to take over the population ($\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_global_condition}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1 .
\end{equation}

If $\alpha<1$, we want to determine a condition for $\tilde{p}'>\tilde{p}$ in \autoref{eq:nextgen_parents_vertical_only}. 
We divide  by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$ to get
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
We divide by $(1-\tilde{p})$ and rearrange so that free terms are on the left and terms with $\tilde{p}$ are on the right. We get
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

Therefore, in this case there can be a third, polymorphic equilibrium, if $0<\tilde{p}^*<1$, where
\begin{equation} \label{eq:vert_hori_equilibrium}
  \tilde{p}^* = 
  \frac{c(1-T_B) - b \alpha T_A - (T_A-T_B)}{b (1-\alpha) (T_A-T_B) }.
\end{equation}

Note that all parameters are positive.
So, applying \autoref{eq:vert_hori_global_condition}, for $\tilde{p}'>\tilde{p}$,  we require that either 
\begin{equation} \label{eq:vert_hori_global_condition1}
T_A>T_B, \;\text{and}\; \tilde{p}>\tilde{p}^*,
\end{equation} or 
\begin{equation} \label{eq:vert_hori_global_condition2}
T_A<T_B, \;\text{and}\; \tilde{p}<\tilde{p}^* .
\end{equation}

We therefore have the following result and corollaries.\\

\begin{result}[Vertical and horizontal transmission of cooperation]
When $0 \le \alpha<1$ and $v=1$, fixation, extinction, and stable coexistence of both cooperation and defection are possible.
\end{result}

Determining the outcome of the dynamics depends on the sign of $T_A-T_B$ and on the legitimacy of the third equilibrium, that is, if $0<\tilde{p}^*<1$.
\\

%This condition can also be formulated as a condition on the cost of cooperation $c$: cooperation will take over the population if
%\begin{equation} \label{eq:unequal_transmission}
%c < \frac{
%	b \big[\alpha T_A + (1-\alpha)\tilde{p}(T_A - T_B)\big] + (T_A - T_B)
%	}{
%	1-T_B
%	}.
%\end{equation}
%
%Importantly, the condition for cooperation to increase from rarity is (set $\tilde{p}=0$ in \autoref{eq:unequal_transmission})
%\begin{equation} \label{eq:unequal_transmission_from_rarity}
%c < \frac{
%	b \alpha T_A + (T_A - T_B)
%	}{
%	1-T_B
%	} .
%\end{equation} 
%
%In general, these conditions cannot be formulated in the form of Hamilton's rule ($c<b\cdot r$) due to the horizontal transmission bias $T_A-T_B$.
%Without horizontal transmission bias, we get the following corollary that does have the form of Hamilton's rule.

\begin{corollary}[Symmetric horizontal transmission]
If $T=T_A=T_B$, then cooperation will take over the population if
\begin{equation}
\label{eq:equal_transmission}
c < b \cdot \frac{\alpha T}{1-T}.
\end{equation}
\end{corollary}
To verify, set $T_A=T_B$ in \autoref{eq:vert_hori_global_condition}.

This can be interpreted as a version of Hamilton's rule (\autoref{eq:hamilton_rule}), where $\alpha T/(1-T)$ is the `effective relatedness'.
\autoref{fig:results_a} demonstrates this condition. 
\\

\begin{corollary}[No assortment of transmission and cooperation]
If $\alpha=0$, then the third equilibrium is
\begin{equation} \label{eq:vert_hori_alpha0_equilibrium}
\tilde{p}^*(\alpha=0) = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B) }.
\end{equation} 
If there is a horizontal transmission bias for cooperation, $T_A>T_B$, then cooperation will fix in the population if the cost is not too high $c < (1+b) \frac{T_A-T_B}{1-T_B}$, and the initial frequency is high enough, $\tilde{p}(0) > \tilde{p}^*$.
If there is a horizontal transmission bias for defection, $T_A<T_B$, defection will take over the population.
\end{corollary}

This can be shown by setting $\alpha=0$ in \autoref{eq:vert_hori_equilibrium}. 
Then, when $T_A>T_B$ we find a condition for $\tilde{p}^*<1$ and $\tilde{p}>\tilde{p}^*$ (\autoref{eq:vert_hori_global_condition1}). When $T_A<T_B$, we have $\tilde{p}^*<0$ and because of \autoref{eq:vert_hori_global_condition2}, any initial frequency leads to fixation of defection, $\tilde{p}=0$.\\

\begin{corollary}[Complete assortment of transmission and cooperation]
\end{corollary}
In this case $\alpha=1$, and horizontal transmission always occurs together with the cooperative interaction; there is complete assortment.
In this case there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found by setting $\tilde{p}'>\tilde{p}$, which gives (\autoref{eq:vert_hori_alpha1})
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
This is equivalent to a result by~\citet[eq.~1]{lewin2017microbes}.

\autoref{eq:vert_hori_alpha1} can be written as
\begin{equation} \label{eq:vert_hori_alpha1_effective}
1 - (1-c)(1-T_B) < (1+b) T_A ,
\end{equation}
which provides an interesting interpretation for the success of cooperation. 
Consider an interaction between two individuals: a cooperator and a defector.
$(1-c)(1-T_B)$ is the probability that the cooperator remains cooperative and also reproduces. 
Therefore, $1 - (1-c)(1-T_B)$ is the probability that either the cooperator becomes a defector, \emph{or} that it fails to reproduce.
This is the effective cost for cooperation from this interaction.
$(1+b) T_A$ is the probability that the defector becomes cooperative and reproduces.
This is the effective benefit for cooperation from this interaction.
So, \autoref{eq:vert_hori_alpha1} means that cooperation can evolve if the effective cost for cooperation is less than the effective benefit.

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation]
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of the cooperative phenotype $A$}, 
\begin{enumerate}
\item if $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$, or
\item if $T_A>T_B$ and $0<\beta_3$, or 
\item if $T_A<T_B$ and $\beta_1<\beta_3$.
\end{enumerate}

@@

\end{enumerate}
\end{result}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation.
We studied fully mixed and very large populations with a prisoner's dilemma payoff. 
We found that under horizontal and vertical cultural transmissions, if \autoref{eq:unequal_transmission} is satisfied, cooperation will take over fully mixed populations (Result 1).
Under oblique and horizontal transmission, horizontal transmission bias for the cooperative phenotype is sufficient and necessary for evolution of cooperation (Result 2, \autoref{eq:oblique_only_result}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are further complicated. Yet, we find that cooperation can evolve and in some cases be maintained together with defection (Result 4). 
Importantly, our results demonstrate that cooperation can evolve even in a fully mixed population (i.e. in an unstructured population), without repeating interactions or individual recognition.
These results significantly further our understating of the cultural evolution of cooperation.

@@

There is an ongoing debate about the extent to which kin selection explains the evolution of cooperation and altruism.
For example, it has been suggested that it can explain the cooperative behavior of worker castes of eusocial insects like the honey bee. % TODO REF
The most significant argument against kin selection is that in some cases cooperation among unrelated individuals appears to have evolved~\citep{wilson2005kin}.
Therefore, other theories have been develop to explain the evolution of cooperation and altruism.

\emph{Reciprocity} entails that repeated interactions or individual recognition are key components of the evolution of cooperation. In \emph{direct reciprocity} there are repeated encounters between the same two individuals, and at every encounter each individual has a choice between cooperation and defection. Hence, it may eventually pay off to cooperate if it may cause your partner to cooperate in the future.
This game-theoretic framework, known as the \emph{repeated prisoner's dilemma}, 
%A well known strategy to play this game is called  'tit-for-tat'. This strategy always starts with a cooperation, then it does whatever the other player has done in the previous round: a cooperation for a cooperation, a defection for a defection. There a unlimited number of possible strategies to play this game. However, 
can only lead to the evolution of cooperation if the cost is less than the benefit $b$ times the probability of another encounter between the same two individuals, $w$, 
%TODO Referencess
\begin{equation} \label{eq:reciprocity}
c < b \cdot w.
\end{equation}

\emph{Group selection} theory posits that cooperation is favored because it imparts an advantage to the whole group, if selection acts at the group level in addition to the individual level. A common model for group selection divides the population into groups in which there are cooperators that help  other group members and defectors that do not.  % TODO add reference
Individuals reproduce proportionally to their fitness, and offspring are added to the same group as their parents.
If a group reaches a certain size it can split to two groups, so groups that grow faster will split more often.
Groups with cooperators grow faster than groups without cooperators, and
cooperation can evolve in this model when the cost $c$ is less than the benefit $b$ times the ratio between the  the number of groups $m$ and the sum of $m$ and the maximum group size $n$,
\begin{equation} \label{eq:groupselection}
c < b \cdot \frac{m}{m+n} .
\end{equation}

\citet{Eshel1982} have shown that with assortative meeting, i.e. probability $m$ that individuals interact with within their phenotypic group, cooperation can evolve if $m > c/b$.
In our Corollary 1.1 (\autoref{eq:equal_transmission}), cooperation evolves if $\alpha T / (1-T) > c/b$. So in our model $\alpha T/(1-T)$ is the effective relatedness, which is affected by $\alpha$, the correlation between transmission and interaction, and $T$, the horizontal transmission rate.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Lilach Hadany and Ayelet Shavit for discussions and comments.
This work was supported in part by
the Israel Science Foundation 552/19 (YR),
and Minerva Stiftung Center for Lab Evolution (YR).
% TODO funding for other authors?
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendices
\begin{appendices}
\renewcommand{\theequation}{\thesection\arabic{equation}}
\counterwithin*{equation}{section}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Bibliography
\bibliographystyle{plainnat}
\bibliography{bib}
